Well-quasi-orderings on word languages

### 2024-10-14 src/infixes.tex
Refactor + todos

@@

\begin{corollary}
    \label{inf-period-union-chains:lem}
    Let $x,u,y \in \Sigma^*$ be words.  The following 
    are finite unions of \kl{chains} for the infix relation:
    $\InfPeriodChain{x}u$, $u \InfPeriodChain{x}$,
    and $\InfPeriodChain{x} u \InfPeriodChain{y}$.
\end{corollary}
\begin{proof}
    Because $\InfPeriodChain{x}$ is a finite union of \kl{chains} for the \kl{suffix}
    relation (rsing \cref{inf-period-chain:lem}), we conclude that $\InfPeriodChain{x}u$ is one too for
    the \kl{infix} relation. This holds similarly for $u \InfPeriodChain{x}$.
    In the case of $\InfPeriodChain{x} u \InfPeriodChain{y}$,
    we remark that $\InfPeriodChain{x} u$ is a finite union of chains
    for the \kl{suffix relation},
    and that $\InfPeriodChain{y}$ is one for the \kl{prefix relation}
    (using again \cref{inf-period-chain:lem}).
    As a consequence,
    $\InfPeriodChain{x}u \InfPeriodChain{y}$ is a finite union 
    of \kl{chains} for the \kl{infix} relation.
\end{proof}

@@

When proving \cref{infix-finite-automata:thm}, we have leveraged a powerful
combinatorial argument on regular langugaes stated in
\cref{pumping-periods:lem}. Our first remark is that this idea can
be generalized to the larger class of \intro{bounded languages}, i.e.,
languages $L \subseteq \Sigma^*$ such that there exists words $w_1, \dots, w_n$
satisfying $L \subseteq w_1^* \cdots w_n^*$.

\begin{theorem}
    \label{bounded-language:thm}
    Let $L$ be a \kl{bounded language} of $\Sigma^*$. Then,
    $L$ is a \kl{well-quasi-order} when endowed with the 
    \kl{infix relation} if and only if it is a finite union of \kl{chains}.
\end{theorem}
\begin{proof}
    Let $w_1, \dots, w_n$ be such that
    $L \subseteq w_1^* \cdots w_n^*$.
    Let us define $m \defined \max \setof{\card{w_i}}{1 \leq i \leq n}$

@@

\subsection{Amalgameation Systems}
\label{infixes-amalgamation:subsec}

@@

    The system $(\Sigma, R, E, \canrun)$ is an \kl{amalgamation system},
    whose language is precisely the language of words recognized
    by the automaton $A$.
\end{example}
\begin{proof}
    By definition, elements of $E(r,s)$ are defined as elements
    of $\HigEmb(\canrun(r), \canrun(s))$, and they compose properly.
    Because $\Delta = Q \times \Sigma \times Q$ is finite, it is 
    a \kl{well-quasi-ordering} when equipped with the equality relation, and 
    we conclude that $\Delta^*$ with $\higleq$ is a \kl{well-quasi-order}
    according to Higman’s Lemma \cite{HIG52}.

@@

\subsection{Downwards Closed Subsets}
\label{infixes-downwards:subsec}

In the light of \cref{reg-wqo-iff-dwclosed-wqo:cor} one may
wonder whether the \kl{downwards closed} languages for the \kl{infix relation}
satisfy a similar characterization in terms of union of \kl{chains} and
\kl{grids} as stated in
\cref{infix-finite-automata:thm,bounded-language:thm,infix-amalgamation:thm}.
Let us first remark that from our previous results we obtain the following
lemma that illustrates how all the previous characterizations of
\kl{well-quasi-ordered} languages for the \kl{infix relation} are collapsing in
the case of \kl{downwards closed} languages.

One may think that all \kl{downwards closed} languages for the \kl{infix
relation} that are \kl{well-quasi-ordered} are \kl{regular}. Note that this is
what happens in the case of the \kl{subword embedding}, where any \kl{downwards
closed} language for this relation is characterized by finitely many excluded
patterns, hence \kl{regular} \textbf{TODO cite}.
However, this is not the case for the \kl{infix relation}, as we will now
illustrate with the following two examples.

\begin{example}
    Let $L \defined a^* b^* \cup b^* a^*$. This language is \kl{downwards
    closed} for the \kl{infix relation}, is \kl{well-quasi-ordered} for the
    \kl{infix relation}, but is characterized by an \emph{infinite} number 
    of excluded infixes of the form $ab^ka$ and $ba^kb$ where $k \geq 1$.
\end{example}

Let $w$ be the Thue-Morse word \textbf{CITE}, and $L$ be the set of infixes of
$w$. By construction $L$ is an \emph{infinite} \kl{downwards closed} for the
\kl{infix relation}. Let us argue that $L$ is \kl{well-quasi-ordered} for the
\kl{infix relation}, but is not \kl{regular}. First, $L$ cannot be \kl{bounded}
because $w$ is \emph{cube free}, that is, no infix of $w$ can be written $uuu$
for some word $u$ \textbf{CITE}. Furthermore, $w$ is a \intro{uniformly
recurrent word} \textbf{CITE}, that is, for every infix $u$ of $w$, there
exists $k \geq 1$ such that $u$ occurs in every infix $v$ of $w$ of size at
least $k$. This implies that $L$ is \kl{well-quasi-ordered} for the \kl{infix
relation}. We conclude thanks to \cref{dwclosed-infixes-wqo:lem},
that $L$ is not \kl{regular}.

>>>>

\todo{Delete this section: the \cref{infix-embedding:thm} is already
    known, we should just restate it. The
    \cref{infix-finite-automata:thm} is a direct consequence of
    \cref{infix-amalgamation:thm} and \cref{bounded-language:thm}.
    I don't know if we should keep it for clarity, but it does not 
    seems to be necessary to me.
}

@@

\begin{corollary}
    \label{inf-period-union-chains:lem}
    Let $x,u,y \in \Sigma^*$ be words.  The following 
    are finite unions of \kl{grids} for the infix relation:
    $\InfPeriodChain{x}u$, $u \InfPeriodChain{x}$,
    and $\InfPeriodChain{x} u \InfPeriodChain{y}$.
\end{corollary}
\begin{proof}
    Because $\InfPeriodChain{x}$ is a finite union of \kl{chains} for the \kl{suffix}
    relation (rsing \cref{inf-period-chain:lem}), we conclude that $\InfPeriodChain{x}u$ is one too for
    the \kl{infix} relation. This holds similarly for $u \InfPeriodChain{x}$.
    In the case of $\InfPeriodChain{x} u \InfPeriodChain{y}$,
    we remark that $\InfPeriodChain{x} u$ is a finite union of chains
    for the \kl{suffix relation},
    and that $\InfPeriodChain{y}$ is one for the \kl{prefix relation}
    (using again \cref{inf-period-chain:lem}).
    As a consequence,
    $\InfPeriodChain{x}u \InfPeriodChain{y}$ is a finite union 
    of \kl{chains} for the \kl{infix} relation.
\end{proof}

@@

When proving \cref{infix-finite-automata:thm}, we have leveraged a powerful
combinatorial argument on regular langugages stated in
\cref{pumping-periods:lem}. Our first remark is that this idea can
be generalized to the larger class of \intro{bounded languages}, i.e.,
languages $L \subseteq \Sigma^*$ such that there exists words $w_1, \dots, w_n$
satisfying $L \subseteq w_1^* \cdots w_n^*$.

\begin{theorem}
    \label{bounded-language:thm}
    Let $L$ be a \kl{bounded language} of $\Sigma^*$. Then,
    $L$ is a \kl{well-quasi-order} when endowed with the 
    \kl{infix relation} if and only if it is a finite union of \kl{grids}.
\end{theorem}
\begin{proof}
    Let $w_1, \dots, w_n$ be such that
    $L \subseteq w_1^* \cdots w_n^*$.
    Let us define $m \defined \max \setof{\card{w_i}}{1 \leq i \leq n}$

@@

Note that this immediately generalizes \cref{reg-wqo-iff-dwclosed-wqo:cor}
to all \kl{bounded languages}.

\begin{corollary}
    \label{bounded-wqo-dwclosed:cor}
    Let $L$ be a \kl{bounded language} of $\Sigma^*$. Then,
    $L$ is a \kl{well-quasi-order} when endowed with the
    \kl{infix relation} if and only if $\dwset[\infleq]{L}$ is.
\end{corollary}

In the light of \cref{reg-wqo-iff-dwclosed-wqo:cor,bounded-wqo-dwclosed:cor}
one may wonder whether the \kl{downwards closed}
languages for the \kl{infix relation} satisfy a similar characterization in
terms of union of \kl{grids} as stated in
\cref{infix-finite-automata:thm,bounded-language:thm}. Let us
first remark that from our previous results we obtain the following lemma that
illustrates how all the previous characterizations of \kl{well-quasi-ordered}
languages for the \kl{infix relation} are collapsing in the case of
\kl{downwards closed} languages.

One may think that all \kl{downwards closed} languages for the \kl{infix
relation} that are \kl{well-quasi-ordered} are \kl{regular}. Note that this is
what happens in the case of the \kl{subword embedding}, where any \kl{downwards
closed} language for this relation is characterized by finitely many excluded
patterns, hence \kl{regular}. However, this is not the case for the \kl{infix
relation}, as we will now illustrate with the following two examples.

\begin{example}
    \label{dwclosed-wqo-not-finite-excl:ex}
    Let $L \defined a^* b^* \cup b^* a^*$. This language is \kl{downwards
    closed} for the \kl{infix relation}, is \kl{well-quasi-ordered} for the
    \kl{infix relation}, but is characterized by an \emph{infinite} number 
    of excluded infixes, respectively of the form $ab^ka$ and $ba^kb$ where $k \geq 1$.
\end{example}

To strengthen \cref{dwclosed-wqo-not-finite-excl:ex}, we will
leverage the \intro{Thue-Morse sequence} $\intro*\ThueMorse \in
\set{0,1}^{\Nat}$, which we will use as a black-box for its two main
characteristics: it is \kl{cube-free} and \kl{uniformly recurrent}. Being
\intro{cube-free} means that no (finite) word of the form $uuu$ is an
\kl{infix} of $\ThueMorse$, and being \intro{uniformly recurrent} means that
for every word $u$ that is an \kl{infix} of $\ThueMorse$, there exists $k \geq
1$ such that $u$ is an \kl{infix} of every word $v$ of size at least $k$. We
refer the reader to a nice survey of Allouche and Shallit for more information
on this sequence and its properties \cite{ALSHA99}.

\begin{lemma}
    \label{thue-morse:lemma}
    There exists a language $L$ that is \kl{downwards closed} for the \kl{infix
    relation}, \kl{well-quasi-ordered} for the \kl{infix relation}, but is not
    \kl{regular}.
\end{lemma}
\begin{proof}
    Let $\ThueMorse$ be the \kl{Thue-Morse sequence}
    $L$ be the set of infixes of $w$. By construction $L$ is an \emph{infinite}
    \kl{downwards closed} for the \kl{infix relation}. Let us argue that $L$ is
    \kl{well-quasi-ordered} for the \kl{infix relation}, but is not \kl{regular}.

    Assume by contradiction that $L$ is \kl{bounded}. In this case, there exist
    words $w_1, \dots, w_k \in \Sigma^*$ such that $L \subseteq w_1^* \cdots
    w_k^*$. Since $L$ is infinite and \kl{downwards closed}, there exists a
    word $u \in L$ such that $u = w_i^3$ for some $1 \leq i \leq k$. This is absurd
    because $u \infleq \ThueMorse$, which is \kl{cube-free}.

    Furthermore, $L$ is \kl{well-quasi-ordered} for the \kl{infix relation}.
    Indeed, consider a sequence $\seqof{u_i}$ of words in $L$. Without loss of
    generality, we may consider a subsequence such that $\card{u_i} <
    \card{u_{i+1}}$ for all $i \in \Nat$. Because $\ThueMorse$ is \kl{uniformly
    recurrent}, there exists $k \geq 1$ such that $u_1$ is an \kl{infix} of
    every word $v$ of size at least $k$. In particular, $u_1$ is an \kl{infix}
    of $u_k$, hence the sequence $\seqof{u_i}$ is \kl(wqo){good}.

    We conclude thanks to
    \cref{dwclosed-infixes-wqo:lem}, that $L$ is not \kl{regular}.
\end{proof}

\subsection{Amalgamation Systems}
\label{infixes-amalgamation:subsec}

@@

    The system $(\Sigma, R, E, \canrun)$ is an \kl{amalgamation system},
    whose language is precisely the language of words recognized
    by the automaton $A$.
\end{example}
\begin{proof}
    By definition, the embeddings inside $E(r,s)$ are defined as elements
    of $\HigEmb(\canrun(r), \canrun(s))$, and they compose properly.
    Because $\Delta = Q \times \Sigma \times Q$ is finite, it is 
    a \kl{well-quasi-ordering} when equipped with the equality relation, and 
    we conclude that $\Delta^*$ with $\higleq$ is a \kl{well-quasi-order}
    according to Higman’s Lemma \cite{HIG52}.

@@

### 2024-10-14 src/introduction.tex
work on intro/contribution (#1)
* work on intro/contribution

* Subfigures for intro graphics

---------

Co-authored-by: lschuetze <lschuetze@mpi-sws.org>
Refactor + todos

\AP One way to measure the complexity of a \kl{well-quasi-ordered} set is to
consider its so-called \intro{ordinal invariants}. Several of these ordinal
measures have been developed for well-quasi-orders. The \intro{maximal order
type} (or m.o.t.) was originally defined by \cite{dejongh77} as the order type
of the maximal linearization of a well-quasi-ordered set. Then,
\cite{schmidt81} introduced the \intro{ordinal height} as the order type of a
maximal chain of a well-quasi-ordered set. Finally, \cite{kriz90b} crafted the
notion of \intro{ordinal width}, standardized the definitions of these three
\kl{ordinal invariants}, and proved numerical relationships between them. For
the \kl{prefix relation}, the \kl{suffix relation}, and the \kl{infix
relation}, the \kl{ordinal height} is always bounded by $\omega$, and the key
parameter to describe its complexity will be the \kl{ordinal width}, i.e.,
``how large the antichains are.''

\todo{rephrase}
\AP In the case of the \kl{prefix relation} and the \kl{suffix relation}, we
conclude from our characterization that the \kl{ordinal width} of a language
$L$ is either finite or $L$ is not \kl{well-quasi-ordered}. The straightforward
generalization of the results for the \kl{prefix relation} and the \kl{suffix
relation} to the \kl{infix relation} is not possible. For example, Kuske has
shown that any countable partial order with finite downward closures can be
embedded into it~\cite{DBLP:journals/ita/Kuske06}. In particular, this means
that the \kl{subword relation} can be embedded into the \kl{infix relation}.
The former being a \kl{well-quasi-ordered} with a very large \kl{ordinal width}
(see for example https://arxiv.org/pdf/2312.14587), this implies that such
"easy characterizations" will not be possible in general. However, not all is
lost: we show that with some mild restrictions we can recover much of the
simple structure of the \kl[prefix relation]{prefix} and \kl{suffix relation}:

@@

We illustrate how this result can be used to study languages recognized by
finite automata, context-free grammars, petri nets, and other systems. This is
possible by leveraging the recently introduced notion of \kl{amalgamation
systems} \cite{ASZZ24}.

>>>>

\AP One way to measure the complexity of a \kl{well-quasi-ordered} set is to
consider its so-called \intro{ordinal invariants}. For instance, the
\intro{order type} of a \kl{well-quasi-ordered} set is the smallest ordinal
that cannot be embedded into it. For the \kl{prefix relation} and the \kl{infix
relation}, we conclude that the order type of a \kl{well-quasi-ordered} set is
at most $k \times \omega$, and its width is finite.\todo{check this + cite}.

\AP The straightforward generalization of the results for the \kl{prefix
relation} and the \kl{suffix relation} to the \kl{infix relation} is not
possible. For example, Kuske has shown that any countable partial order with
finite downward closures can be embedded into
it~\cite{DBLP:journals/ita/Kuske06}. In particular, this means that the
\kl{subword relation} can be embedded into the \kl{infix relation}. The former
being a \kl{well-quasi-ordered} with \kl{order type} $\omega^\omega$
\todo{cite}. However, not all is lost: we show that with some mild restrictions
we can recover much of the simple structure of the \kl[prefix relation]{prefix}
and \kl{suffix relation}:

@@

We illustrate how this result can be used to study languages recognized by
finite automata, context-free grammars, petri nets, and other systems. This is
possible by leveraging the recently introduced notion of \kl{amalgamation
systems} \cite{ASZZ24}, which abstracts the notion of systems whose runs can be
suitably ``pumped'' and include common language classes such as regular
languages, context-free languages, and even languages recognized by petri nets.
